---
output: rmarkdown::github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file    -->

<!-- Origin: https://github.com/hrbrmstr/ggalt/blob/master/README.Rmd -->
<!-- Thanks to Bob Rudis for sharing.                                 -->
```{r, echo = FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse=TRUE, 
                      comment="##", 
                      fig.retina=2, 
                      fig.path = "README_figs/README-")
```

# Monty Hall Problem - Vectorised Simulations in R

![Monty Hall Problem Image](README_files/Monty_open_door.svg)

This repository presents three vectorised simulations to the "Monty Hall Problem" 
in [R](https://r-project.org) using:

1. Base R.
2. A [tinyverse](http://tinyverse.org/) approach using the `data.table` package.
3. A [tidyverse](https://tidyverse.org) approach using `dplyr` and `tidyr` packages.

## Monty Hall Problem

There is a good write up and discussion about the Monty Hall Problem on Wikipedia from 
which I quote:

*The Monty Hall problem is a brain teaser, in the form of a probability puzzle, loosely based on the American television game show Let's Make a Deal and named after its original host, Monty Hall.* 

**Question**

> Suppose you're on a game show, and you're given the choice of three doors: Behind one door is a car; behind the others, goats. You pick a door, say No. 1, and the host, who knows what's behind the doors, opens another door, say No. 3, which has a goat. He then says to you, "Do you want to pick door No. 2?" Is it to your advantage to switch your choice? 

**Standard Assumptions**

+ _The host must always open a door that was not picked by the contestant._
+ _The host must always open a door to reveal a goat and never the car._
+ _The host must always offer the chance to switch between the originally chosen door and the remaining closed door._

**Source:** https://en.Wikipedia.org/wiki/Monty_Hall_problem

## Objective

> Use a vectorised approach to simulate the probability of winning for both, staying with 
the chosen door and switching door when offered.

## Simulations 

The idea is to simulate multiple games (and outcomes) to evaluate whether staying or 
switching would be better. 

We set the following parameters to make these simulations reproducible :
```{r simSetup}
set.seed(663948)
simNum <- 100000  # Number of simulations
doorNum <- 3      # Number of doors (not strictly necessary)
```

Each of the simulations below will create the following variables:

+ `sim` -- simulation number.
+ `door` -- door number.
+ `true` -- behind true door: 1 = car & 0 = goat.
+ `guess` -- contestant's chosen door: 1 = car & 0 = goat.
+ `switch` -- door offered to the contestant to switch to.

These are then summarised as follows:

+ `stayWin`  -- proportion of winning by staying with `guess` door.
+ `switchWin`  -- proportion of winning by switching to `switch` door.

## Base R (R Zeroverse) Simulation

The first simulation only uses Base R functions. I call it "R zeroverse" as CRAN 
contributed packages are not allowed, just those that come with R by default (e.g. 
`base`, `graphics` `stats`).


```{r brSim}
# A function to simulate randomly chosen doors for the truth door and the guess.
rDoors <- function(sims){
  unlist(lapply(1:sims, function(x) sample(c(0, 0, 1), replace = FALSE)))
}

# Note that for this method to work the data must be sorted by "sim" then the door 
# (hence the order in the expand.grid() function).
mhbrGrid <- expand.grid(door=1:doorNum, sim=1:simNum)
mhbrGrid$true <- rDoors(simNum)
mhbrGrid$guess <- rDoors(simNum)

mhbr <- mhbrGrid[with(mhbrGrid, order(sim, -guess, true)), ]
mhbr$switch <- rep(c(0, 0, 1), simNum)
mhbr$stayWin <- with(mhbr, guess*true)
mhbr$switchWin <- with(mhbr, switch*true)

mhBaseR <- rbind(list(stayWin = sum(mhbr$stayWin)/simNum, 
                      switchWin = sum(mhbr$switchWin)/simNum))
mhBaseR
```

## Tinyverse (`data.table`) Simulation

A neat [tinyverse](http://tinyverse.org/) simulation using [`data.table`](http://r-datatable.com/)
package.

```{r dtSim, message=FALSE}
library(data.table)

simdt <- CJ(sim=1:simNum, door=1:doorNum)

mhdt <-  simdt[, `:=`(guess=sample(c(0, 0, 1)), true=sample(c(0, 0, 1)), switch=0), .(sim)][
    order(sim, -guess, true)][
    rowid(sim)==doorNum, switch:=1][
    , .(stayWin = sum(guess*true)/simNum, switchWin = sum(switch*true)/simNum)]

mhdt
```

## Tidyverse (`dplyr` & `tidyr`) Simulation

A neat [tidyverse](https://tidyverse.org) simulation using `dplyr` and `tidyr` packages.

```{r tvSim, message=FALSE}
library(dplyr)
library(tidyr)

mhtv <- expand_grid(sim=1:simNum, door=1:doorNum) %>%
    group_by(sim) %>%
    mutate(guess=sample(c(0, 0, 1)), true=sample(c(0, 0, 1))) %>%
    arrange(sim, -guess, true) %>%
    mutate(switch = as.numeric(row_number(sim)==doorNum)) %>%
    ungroup %>%
    summarise(stayWin = sum(guess*true)/simNum,
              switchWin = sum(switch*true)/simNum)

mhtv
```

## Summary

+ All three simulations lead to the same conclusion -- that probabilistically it is
better to switch. 
+ I provide three vectorised solutions as the R community is diverse and each member/group
has different preferences. 
    + There are other ways to simulate this problem.
    + Using a `for` loop based approach is perfectly fine. 
    + I just wanted code it using a vectorised approach as it was bugging me ;) 

**Thanks for reading.**

## Acknowledgements

Thanks to the [R core](https://www.r-project.org/contributors.html), [`data.table`](https://rdatatable.gitlab.io/data.table/authors.html) and
[tidyverse](https://tidyverse.org) authors, maintainers and contributors.
